Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-21 tex/classical.tex
Rewording: fist part of sec. 2.3.2

The distribution functions, $\{f_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Each $f_M$ can be formally extended to the complete coordinate space
by defining each $f_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f_M(\vec v,t)\}$:
%
\begin{equation}
   f(\vec r,\vec v,t)=\sum_{M=1}^Kf(\vec r_M,\vec v,t)=\sum_{M=1}^Kf_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    H_M' & = &  \int f_M(\vec{v},t) \ln f_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are surmising the 
local equilibrium hypothesis. In addition, the set $\{f_{M}(\vec{v},t)\}$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f_M$
does not depend on the cell number $M$ (\textit{i.e.} $f_M(\vec v,t)=f(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    H'(t)=\int \sum_{M=1}^{K} [f(\vec{v},t)\ln f(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $H'$ with the entropy, it remains to show that $H'(t)$ is extensive,
with respect to $f'(\vec v,t)\big(\equiv Kf_(\vec v,t)\big)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int Kf(\vec{v},t) \ln \big[Kf(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f(\vec{v},t) + Kf(\vec{v},t) \ln f(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f(\vec{v},t)\big[\ln K + \ln f(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v \simeq
    K\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $H_M'$ is extensive, and the sum $\sum_MH'_M$ is the $H$-functional of the complete
system and reduces to the Boltzmann $H$-functional. Hence, $H'$ can also be identified with the
system's entropy.

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f_M(\vec{v})$
reduces to
%
\begin{equation}
    f_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $H''$
functional:
%
\begin{equation}
   H''(t)=\sum_{M=1}^{K}\int \left[f_M(\vec{v},t)
    \ln f_M(\vec{v},t)-f_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
however $\sum_{M=1}^{K} \int f_M(\vec{v},t)\mathrm{d}^3v =N$.
Since $N$ is a constant and we are mainly intersted in the time-derivative of $H''$,
$C$ can be conveniently omitted. In other words, $H'$ renders the Maxwell-Boltzmann
distribution function for systems in equilibrium.

\subsubsection{Alternative $H$-theorem}

To prove the Boltzmann's $H$-theorem (using (\ref{eq:CH2}) as functional), we need
to assume that the distribution function $f_M(\vec{v},t)$ must satisfy the
following assumptions:
%
\begin{itemize}
  \item The local equilibrium hypothesis in each cell ($f_{M}$ must be
    spatial-homogeneous for each $M$).
  \item The non-homogeneous distribution function assumption in the total
    volume ($f_{M}$ must be different among all cells).
  \item The following restrictions
    %
    \begin{equation}\label{restrictionoutsideclassical}
      \int f_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}+\Delta_M(t); \ \ \ \ 
      \int f_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}+ \delta_M(t),
    \end{equation}
    %
    where $\bar {\mathcal{N}}$ and $\bar{\mathcal{E}}$ are
    the local particle number and the local energy in equilibrium
    %
    \begin{equation}
      \bar{\mathcal{N}}=
      \int \bar{f}(\vec{v}) \mathrm{d}^{3}v ; \quad
      \bar{\mathcal{E}}=
      \int \bar{f}(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v,
    \end{equation}
    %
    and $\Delta_M$, $\delta_M$ could be seen as a deviation from
    $\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$ respectively,
    with $\Delta_M(t)\ll\bar{\mathcal{N}}$ and
    $\delta_M(t)\ll\bar{\mathcal{E}}$.
\end{itemize} 
%
Those conditions described a system out of equilibrium but no so far from it,
that is a perturbed system and will suffer a \textit{relaxation process, i.e.}
The time evolution of a perturbated system to the equilibrium state.

It is important to remark that the quantities $\Delta_M$, $\delta_M$ are
sufficiently big to be different from fluctuations in the system but too small
such that the system is not far from the equilibrium state.

Performing the derivative of (\ref{eq:CH2}) with respect to time, we obtain
%
\begin{equation}\label{dH1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f_M(\vec{v},t)
    \right]\dot f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Using the first-order approximation
%
\begin{equation}\label{firstorder}
   f_{Mn}(\vec{v},t)=\bar{f}_{n}(\vec{v})(1+g_{Mn}(\vec{v},t)): \ \ \ 
   \bar{f}_{n}(\vec{v})\gg \bar{f}_{n}(\vec{v})|g_{Mn}(\vec{v},t)|,
\end{equation}
%
(\ref{dH1}) yields
%
\begin{equation}\label{dH1-1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\bar f(\vec{v}) \left[
      1+\ln \left\{
        \bar f(\vec{v})+\bar f(\vec{v})g_M(\vec{v},t)
      \right\}
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

>>>>

The distribution functions, $\{f_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Each $f_M$ can be formally extended to the complete coordinate space
by defining each $f_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f_M(\vec v,t)\}$:
%
\begin{equation}
   f(\vec r,\vec v,t)=\sum_{M=1}^Kf(\vec r_M,\vec v,t)=\sum_{M=1}^Kf_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f_M(\vec v,t)\neq f_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    H_M' & = &  \int f_M(\vec{v},t) \ln f_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are surmising the 
local equilibrium hypothesis. In addition, the set $\{f_{M}(\vec{v},t)\}$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f_M$
does not depend on the cell number $M$ (\textit{i.e.} $f_M(\vec v,t)=f(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    H'(t)=\int \sum_{M=1}^{K} [f(\vec{v},t)\ln f(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $H'$ with the entropy, it remains to show that $H'(t)$ is extensive,
with respect to $f'(\vec v,t)\big(\equiv Kf_(\vec v,t)\big)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int Kf(\vec{v},t) \ln \big[Kf(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f(\vec{v},t) + Kf(\vec{v},t) \ln f(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f(\vec{v},t)\big[\ln K + \ln f(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v \simeq
    K\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $H_M'$ is extensive, and the sum $\sum_MH'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $H'$ can also be identified with the
system's entropy.

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $H''$
functional:
%
\begin{equation}
   H''(t)=\sum_{M=1}^{K}\int \left[f_M(\vec{v},t)
    \ln f_M(\vec{v},t)-f_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
however $\sum_{M=1}^{K} \int f_M(\vec{v},t)\mathrm{d}^3v =N$.
Since $N$ is a constant and we are mainly intersted in the time-derivative of $H''$,
$C$ can be conveniently omitted. In other words, $H'$ renders the Maxwell-Boltzmann
distribution function for systems in equilibrium.

\subsubsection{Alternative proof of the classical $H$-theorem}

Throughout this section, we will consider that the classical gas
starts at some time as an out-of-equilibrium system,
but in such a manner that it is inteuitively not too far from equilibrium,
which ensures that the local equilibrium hypothesis remains valid during
the time evolution of the system.
Also we will use the following definitions:
\begin{subequations}\label{eq:restrictionoutsideclassical}
%
\begin{equation}
      \mathcal{N}'_M(t)=\int f'_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}'+\Delta'_M(t)
\end{equation}
%
and
%
\begin{equation}
      \mathcal{E}'_M(t)=\int f'_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}'+ \delta'_M(t).
\end{equation}
\end{subequations}
%
Here $\bar {\mathcal{N}}'\ (=N/K)$ and $\bar{\mathcal{E}}'\ (=E/K)$ are
the local particle number and the local energy in equilibrium, respectively, which are given by:
%
    \begin{equation}
      \bar{\mathcal{N}}'=
      \int \bar{f}(\vec{v}) \mathrm{d}^{3}v\quad\textrm{and} \quad
      \bar{\mathcal{E}}'=
      \int \bar{f}(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v.
    \end{equation}
    %
Here, we have used $\bar{f}(\vec{v})$ as defined in Eq.~(\ref{eq:fbardef}).
In Eqs.~(\ref{eq:restrictionoutsideclassical}), $\Delta'_M$ and $\delta'_M$
are considered deviations relative to
$\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$, respectively, and for systems that are not
too far away from equilbrium, it is reasonable to expect on the one hand that
$\Delta'_M(t)\ll\bar{\mathcal{N}}'$ and
$\delta'_M(t)\ll\bar{\mathcal{E}}'$, and on the other hand that $\Delta'_M$ and
$\delta'_M$ are sufficiently large compared with fluctuations of $\bar{\mathcal{N}}'$
and $\bar{\mathcal{E}}'$.

\commr{CHECKPOINT}\\
Performing the derivative of (\ref{eq:CH2}) with respect to time, we obtain
%
\begin{equation}\label{dH1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f_M(\vec{v},t)
    \right]\dot f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Using the first-order approximation
%
\begin{equation}\label{firstorder}
   f_{Mn}(\vec{v},t)=\bar{f}_{n}(\vec{v})(1+g_{Mn}(\vec{v},t)): \ \ \ 
   \bar{f}_{n}(\vec{v})\gg \bar{f}_{n}(\vec{v})|g_{Mn}(\vec{v},t)|,
\end{equation}
%
(\ref{dH1}) yields
%
\begin{equation}\label{dH1-1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\bar f(\vec{v}) \left[
      1+\ln \left\{
        \bar f(\vec{v})+\bar f(\vec{v})g_M(\vec{v},t)
      \right\}
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
